A Framework for Reproducible Data Science using Nix

### 2025-10-20 paper.qmd
updated example

The user defines the pipeline in an [R]{.proglang} script using functions
inspired by [targets]{.pkg}. For example, a simple pipeline might read data with
[Python]{.proglang} (`polars`), filter it, process it with [R]{.proglang}
(`dplyr`), and compile a Quarto document. The pipeline is defined as a list,
typically in a file called `gen-pipeline.R` and its execution environment in a
file called `gen-env.R`. Templates of these files can be created using
`rxp_init()`. On a machine that has [Nix]{.pkg} installed it is possible to
bootstrap an environment using a temporary shell:

```{=tex}
\begin{CodeInput}
$> nix-shell -I \
  nixpkgs=https://github.com/rstats-on-nix/nixpkgs/archive/refs/heads/2025-10-17.tar.gz -p \
  R rPackages.rix rPackages.rixpress
\end{CodeInput}
```
One can start the R interpreter, and then call `rxp_init()`:

```{=tex}
\begin{CodeInput}
R> rixpress::rxp_init()
\end{CodeInput}
```

Let us modify `gen-pipeline.R`:

@@

R> pipeline_steps <- list(
+  # Step 1: Read data using Python
+  rxp_py_file(
+    name = mtcars_pl,
+    path = "data/mtcars.csv",
+    read_function = "lambda x: polars.read_csv(x, separator = '|')"
+  ),
+
+  # Step 2: Filter the data in Python
+  rxp_py(
+    name = mtcars_filtered,
+    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()"
+  ),
+
+  # Step 3: Convert Python pandas DataFrame to R data.frame
+  # This uses reticulate for conversion. It would also be possible
+  # to pass objects to and from languages by serializing them
+  # to json for instance, or another universal format
+  rxp_py2r(
+    name = mtcars_r,
+    expr = mtcars_filtered
+  ),
+
+  # Step 4: Process data in R
+  rxp_r(
+    name = mtcars_summary,
+    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
+  )
+)

# Generate the 'pipeline.nix' file from the R list
R> rxp_populate(pipeline_steps)
\end{CodeInput}
```

The required execution environment, defined in `gen-env.R`, is as follows:

```{=tex}
\begin{CodeInput}
R> library(rix)

R> rix(
+   date = "2025-10-14",
+   r_pkgs = c("dplyr", "ggdag", "quarto", "reticulate", "rixpress"),
+   py_conf = list(
+     py_version = "3.13",
+     py_pkgs = c("polars", "pyarrow", "pandas")
+   ),
+   ide = "none",
+   project_path = ".",
+   overwrite = TRUE
+ )
\end{CodeInput}
```

Using the same temporary shell, users can now generate the `default.nix` file
by calling `source("gen-env.R")`. Users can now quit the temporary shell, and
build the environment by calling `nix-shell` in a terminal. Once the
development (or execution) environment has been built, users can enter it,
and call `source("gen-pipeline.R")` to generate a file called `pipeline.nix`
which is the [Nix]{.pkg} expression that defines the entire reproducible
pipeline.

The environment also includes the [ggdag]{.pkg} package [@ggdag], so it's
possible to view a visual representation of the pipeline (yellow nodes are
[Python]{.pkg} and blue nodes are [R]{.pkg} derivations respectively):

![Graphical representation of the pipeline](dag.png)

Running the script above does not execute the pipeline; it generates a
`pipeline.nix` file, which declaratively defines the workflow.

To build and execute the pipeline:

```{=tex}
\begin{CodeInput}
R> rxp_make()
\end{CodeInput}
```

```{=tex}
\begin{CodeInput}
Build process started...

+ mtcars_pl building
+ mtcars_filtered building
+ mtcars_r building
+ mtcars_summary building
✓ mtcars_filtered built
✓ mtcars_pl built
✓ mtcars_r built
✓ mtcars_summary built
✓ pipeline completed [4 completed, 0 errored]
Build successful! Run `rxp_inspect()` for a summary.
Use `rxp_read("derivation_name")` to read objects or
`rxp_load("derivation_name")` to load them into the global environment.
\end{CodeInput}
```

[Nix]{.pkg} executes each step in order, building dependencies as needed.
Outputs are cached, so subsequent runs only recompute steps with changed inputs
or code. The build process can also be executed in parallel, if the the pipeline
has independent branches. This provides the efficiency of [targets]{.pkg} with
multi-language support and bit-for-bit reproducibility. Artifacts can be
inspected interactively in [R]{.proglang} using `rxp_read("artifact_name")` or
`rxp_load("artifact_name")`.

For Python users, a port called [ryxpress]{.pkg} allows building the same
pipelines and inspecting outputs from Python sessions.

[rixpress]{.pkg} includes several additional features not covered here for the
sake of brevity.

It is also possible to configure popular IDEs to work interactively and
seamlessly with both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth,
reproducible workflow from within the development environment. Detailed setup
instructions are provided in the vignettes of both packages.

@@

>>>>

To illustrate these capabilities with a realistic research task, we present a
polyglot pipeline that simulates a Real Business Cycle (RBC) model in
[Julia]{.proglang}, uses the resulting data to train an [XGBoost]{.pkg}
forecasting model in [Python]{.proglang}, visualizes the results in
[R]{.proglang} with [ggplot2]{.pkg}, and finally compiles a [Quarto]{.pkg}
report.

The user defines the pipeline in an [R]{.proglang} script (`gen-pipeline.R`)
using functions inspired by [targets]{.pkg}. The underlying logic for each step
is encapsulated in separate helper scripts (`functions.jl`, `functions.py`, and
`functions.R`), the full contents of which are detailed in the Appendix. The
high-level orchestration script demonstrates how `{rixpress}` defines a
granular, multi-step machine learning workflow that crosses language boundaries:

@@

R> pipeline_steps <- list(
+  # STEP 0: Define RBC Model Parameters in Julia
+  rxp_jl(alpha, 0.3),
+  rxp_jl(beta, 1 / 1.01),
+  # ... (other parameters omitted for brevity) ...
+
+  # STEP 1: Julia - Simulate the RBC model
+  rxp_jl(
+    name = simulated_rbc_data,
+    expr = "simulate_rbc_model(alpha, beta, delta, rho, sigma, sigma_z)",
+    user_functions = "functions/functions.jl",
+    encoder = "arrow_write"
+  ),
+
+  # STEP 2.1: Python - Prepare features
+  rxp_py(
+    name = processed_data,
+    expr = "prepare_features(simulated_rbc_data)",
+    user_functions = "functions/functions.py",
+    decoder = "pyarrow.feather.read_feather"
+  ),
+
+  # STEP 2.2: Python - Split data (X_train, y_train, etc.)
+  rxp_py(name = X_train, expr = "get_X_train(processed_data)", ...),
+  # ... (other data splits omitted for brevity) ...
+
+  # STEP 2.3: Python - Train the XGBoost model
+  rxp_py(
+    name = trained_model,
+    expr = "train_model(X_train, y_train)",
+    user_functions = "functions/functions.py"
+  ),
+
+  # STEP 2.4: Python - Make predictions and format results
+  # ... (prediction and formatting steps omitted for brevity) ...
+  rxp_py(
+    name = final_predictions_df,
+    expr = "format_results(y_test, model_predictions)",
+    user_functions = "functions/functions.py",
+    encoder = "save_arrow"
+  ),
+
+  # STEP 3: R - Visualize the predictions
+  rxp_r(
+    name = output_plot,
+    expr = plot_predictions(final_predictions_df),
+    user_functions = "functions/functions.R",
+    decoder = arrow::read_feather
+  ),
+
+  # STEP 4: Quarto - Compile the final report
+  rxp_qmd(
+    name = final_report,
+    qmd_file = "readme.qmd"
+  )
+)

# Generate the 'pipeline.nix' file from the R list
R> rxp_populate(pipeline_steps, build = TRUE)
\end{CodeInput}
```

The `rxp_populate()` function translates this [R]{.proglang} list into a
`pipeline.nix` file, which declaratively defines the entire workflow. The
package can then generate a visual representation of the pipeline's directed
acyclic graph (DAG), as shown in Figure 1.

![Graphical representation of the polyglot pipeline. Green nodes are
[Julia]{.proglang}, yellow nodes are [Python]{.pkg}, and blue nodes are
[R]{.proglang} derivations.](dag-polyglot.png){#fig-dag}

To execute the pipeline, one can either set `build = TRUE` in `rxp_populate()`
or call `rxp_make()` separately. [Nix]{.pkg} executes each step in order,
building dependencies as needed. Outputs are cached, so subsequent runs only
recompute steps with changed inputs or code. This provides the efficiency of
[targets]{.pkg} with multi-language support and bit-for-bit reproducibility.
Artifacts can be inspected interactively in [R]{.proglang} using
`rxp_read("artifact_name")` or `rxp_load("artifact_name")`.

For Python users, a port called [ryxpress]{.pkg} allows building the same
pipelines and inspecting outputs from Python sessions. [rixpress]{.pkg} also
includes several additional features not covered here for brevity. It is also
possible to configure popular IDEs to work interactively and seamlessly with
both [rix]{.pkg} and [rixpress]{.pkg}, enabling a smooth, reproducible workflow
from within the development environment. Detailed setup instructions are
provided in the vignettes of both packages.

@@

## Appendix: A Complete Polyglot Example with rixpress {#sec-appendix}

This appendix provides the complete source code for the polyglot pipeline
example discussed in Section 5. The pipeline simulates a Real Business Cycle
(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
[Python]{.proglang}, and visualizes the results in [R]{.proglang}.

### The Environment Definition (`gen-env.R`)

This script uses [rix]{.pkg} to define a `default.nix` file that contains all
necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
[Python]{.proglang}.

```{=tex}
\begin{CodeInput}
R> # This script defines the polyglot environment our pipeline will run in.
R> library(rix)
R>
R> # Define the complete execution environment
R> rix(
+   # Pin the environment to a specific date for reproducibility.
+   date = "2025-10-14",
+
+   # R Packages
+   r_pkgs = c(
+     "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
+   ),
+ 
+   # Julia Configuration
+   jl_conf = list(
+     jl_version = "lts",
+     jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
+   ),
+ 
+   # Python Configuration
+   py_conf = list(
+     py_version = "3.13",
+     py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
+   ),
+ 
+   project_path = ".",
+   overwrite = TRUE
+ )
\end{CodeInput}
```

### The RBC Model Simulation (`functions/functions.jl`)

This Julia script implements the state-space solution to the RBC model as a pure
function. The economic theory is based on @depaoli2009.

```{=tex}
\begin{Code}
# This script contains the helper functions for the Julia portion of the pipeline.
# It implements the correct, stable, and canonical log-linearized solution to
# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.

# In an interactive session, you should uncomment the line below
# to load the required packages. rixpress handles this automatically.
#using LinearAlgebra, Distributions, DataFrames, Arrow, Random

# 1. Main Simulation Function

"""
    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    Takes RBC model parameters as input, solves for the state-space representation
    using the method of undetermined coefficients (as per the slides), simulates
    the model for 250 quarters, and returns a DataFrame.
"""
function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
    y_k = ((1/β) - 1 + δ) / α

    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
    # We solve for the coefficients of the policy functions:
    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t

    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
    # Based on Campbell (1994) and slide 22, we take the stable root.
    # a*x^2 + b*x + c = 0
    a_quad = y_k - δ
    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
    c_quad = y_k * α

    # The stable solution is the smaller positive root
    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)

    # Now solve for the other coefficients based on c_ck
    c_kk = (y_k * α) / (y_k - δ - c_ck)
    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))

    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
    # State vector: s_t = [k_{t-1}, z_t]'
    # Transition:   s_t = T*s_{t-1} + R*ε_t

    T = [c_kk  c_kz * ρ
         0     ρ      ]

    R = [c_kz ; 1]

    # Observation Matrix for [output, consumption, investment]
    # y_hat_t = α*k_{t-1} + z_t
    # c_hat_t = c_ck*k_{t-1} + c_cz*z_t
    # i_hat_t = (i_y_ss)^-1 * (k_t - (1-δ)k_{t-1})
    #         = (i_y_ss)^-1 * ( (c_kk - (1-δ))k_{t-1} + c_kz*z_t )

    i_y_ss = δ * (α / ((1/β) - 1 + δ))

    C = [α      1
         c_ck   c_cz
         (c_kk - (1-δ))/i_y_ss   c_kz/i_y_ss]

    # --- SIMULATE THE MODEL ---
    Random.seed!(1234)
    n_periods = 250
    shocks = randn(n_periods) * σ_z

    states = zeros(2, n_periods)
    for t in 2:n_periods
        # The state is [k_{t-1}, z_t]'
        # To get the next state [k_t, z_{t+1}]', we first find k_t
        k_t = T[1,1]*states[1, t] + T[1,2]*states[2, t] + R[1]*shocks[t]
        z_tp1 = T[2,1]*states[1, t] + T[2,2]*states[2, t] + R[2]*shocks[t]

        states[1, t] = k_t # This is now k_t, which is "k_lag" for t+1
        states[2, t] = z_tp1 # This is now z_{t+1}
    end
    # Re-aligning states to be [k_{t-1}, z_t]
    k_lag = [0; states[1, 1:end-1]]
    z = [0; states[2, 1:end-1]]

    observables = C * [k_lag'; z']

    df = DataFrame(
        period = 1:n_periods,
        output = observables[1, :],
        consumption = observables[2, :],
        investment = observables[3, :],
        capital = k_lag,
        technology = z
    )

    return df
end

# 2. Encoder Function (Unchanged)
"""
    arrow_write(df::DataFrame, path::String)
    Encoder function to save a Julia DataFrame to an Arrow file.
"""
function arrow_write(df::DataFrame, path::String)
    Arrow.write(path, df)
end
\end{Code}
```

### The XGBoost Forecasting Model (`functions/functions.py`)

This Python script contains a series of modular functions to handle feature
engineering, data splitting, model training, and prediction.

```{=tex}
\begin{Code}
# This script contains modular helper functions for the Python portion of the pipeline.
# Each function performs a single, distinct task in the ML workflow.

# 1. Load required Python packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 to 11.
# With rixpress, the packages get loaded automatically.
#import pandas as pd
#import pyarrow.feather as feather
#from sklearn.model_selection import train_test_split
#import xgboost as xgb
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.

# Step 1: Feature Engineering
def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
    """Takes the raw simulated data and creates lagged features and the target variable."""
    df = simulated_df.copy()

    # Create lagged features
    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
        df[f'{col}_lag1'] = df[col].shift(1)

    # Drop the first row which now has NaNs
    df.dropna(inplace=True)

    return df

# Step 2: Data Splitting Functions
def get_X_train(processed_df: pd.DataFrame):
    """Gets the training features (X_train) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[:train_size]

def get_y_train(processed_df: pd.DataFrame):
    """Gets the training target (y_train) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[:train_size]

def get_X_test(processed_df: pd.DataFrame):
    """Gets the testing features (X_test) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[train_size:]

def get_y_test(processed_df: pd.DataFrame):
    """Gets the testing target (y_test) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[train_size:]

# Step 3: Model Training
def train_model(X_train: pd.DataFrame, y_train: pd.Series):
    """Initializes and trains the XGBoost model."""
    model = xgb.XGBRegressor(
        objective='reg:squarederror',
        n_estimators=100,
        learning_rate=0.1,
        max_depth=3,
        random_state=42
    )
    model.fit(X_train, y_train)
    return model

# Step 4: Prediction
def make_predictions(model, X_test: pd.DataFrame):
    """Makes predictions on the test set using the trained model."""
    return model.predict(X_test)

# Step 5: Format Final Results
def format_results(y_test: pd.Series, predictions) -> pd.DataFrame:
    """Combines test data and predictions into a final DataFrame for R."""
    results_df = pd.DataFrame({
        'period': y_test.index,
        'actual_output': y_test.values,
        'predicted_output': predictions
    })
    return results_df

# Encoder Function (for saving the final output)
def save_arrow(df: pd.DataFrame, path: str):
    """Encoder function to save a pandas DataFrame to an Arrow file."""
    feather.write_feather(df, path)

\end{Code}
```

### The Visualization (`functions/functions.R`)

This R script uses [ggplot2]{.pkg} to create a plot comparing the model's output
against the XGBoost predictions.

```{=tex}
\begin{CodeInput}
# This script contains the helper functions for the R portion of the pipeline.
# It defines the visualization logic using ggplot2.

# 1. Load required R packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 and 9.
# With rixpress, the packages get loaded automatically.
#library(ggplot2)
#library(dplyr)

#-------------------------------------------------------------------------------
# 2. Main Function: Create the Visualization
#-------------------------------------------------------------------------------

#' Create a plot comparing actual vs. predicted output
#'
#' This function takes a data frame with actual and predicted time series data
#' and generates a ggplot visualization.
#'
#' @param predictions_df A data frame containing columns: 'period',
#'   'actual_output', and 'predicted_output'. This data frame will be the
#'   output of the Python XGBoost step.
#' @return A ggplot object.
#'
plot_predictions <- function(predictions_df) {
  # Create the plot object
  p <- ggplot(predictions_df, aes(x = period)) +

    # Add a line for the actual output from the RBC model simulation
    geom_line(
      aes(y = actual_output, color = "Actual (RBC Model)"),
      linewidth = 1
    ) +

    # Add a dashed line for the XGBoost model's predictions
    geom_line(
      aes(y = predicted_output, color = "Predicted (XGBoost)"),
      linetype = "dashed",
      linewidth = 1
    ) +

    # Define custom colors and legend labels
    scale_color_manual(
      name = "Series",
      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
    ) +

    # Add informative labels and a title
    labs(
      title = "XGBoost Prediction of RBC Model Output",
      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
      x = "Time (Quarters)",
      y = "Output (Log-deviations from steady state)"
    ) +

    # Use a clean theme
    theme_minimal() +

    # Improve legend position
    theme(legend.position = "bottom")

  # Return the final ggplot object
  return(p)
}
\end{CodeInput}
```
